Automatic Identification and Annotation of Parallel Tasks in Structured Programs

### 2018-01-18 docs/ACM/paper.tex
toys tests updated. added quadtree.
Section 2.

\section{Advantages and Difficulties}
\label{sec:ovf}

The use of OpenMP pragmas as the concrete means to explore, without human
intervention, task parallelism in programs presents difficulties and
advantages. 
Such challenges and benefits have guided the design and implementation of the
techniques described in Section~\ref{sec:sol}.
In the rest of this section we discuss such advantage and difficulties.

\subsection{The Advantages}
\label{sub:adv}

The main advantage of using OpenMP annotations to split a program into parallel
tasks is
the ability  of leaving the job of  handling dependences to a runtime.
The OpenMP runtime maintains a task dependence graph which dynamically exposes more parallelism 
opportunities than those resulting from a conservative  compile-time  analysis.
In particular, the runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its correctness~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}. \ed{To achieve that, OpenMP runtime uses task descriptors that contain}
\guido{The descriptor does not keep task status; this is done in the runtime.}
\rmv{task's status (running, waiting)} \ed{a task identifier and  a list of memory addresses corresponding to the task input /output arguments.}

The process of checking for dependencies can be as simple as checking if an
incoming task accesses a memory address already in use by \ed{an in-flight task}.
It can also be complex, involving checks over long ranges of addresses
(in case the runtime system supports dependence checking for memory ranges).
This process can also include memory labelling and renaming (in case the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a \new{dynamic} topological ordering of this
graph~\cite{Planas15}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\columnwidth]{images/ex_book_filter}
\caption{The benefits of the OpenMP runtime environment.
Neither the runtime checks of Alves~\cite{Alves15} or Rus~\cite{Rus02}, nor
Whaley's context and flow sensitive alias analysis would be able to ensure the
correctness of the automatic parallelization of this program.}
\label{fig:ex_book_filter}
\end{center}
\end{figure}

\guido{Now it is clear the relevance for irregular programs.}
\ed{The OpenMP runtime} support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
\guido{A litte bit more details: }
\new{In such programs, control construcs (e.g. {\tt if} statement) make the execution 
of certain pieces of code  dependent on the program input. When using 
task parallelism the runtime can capture such dependences. This is not possible 
otherwise, since compile-time dependence analyzers must be conservative and eventually
skip some parallelization opportunity.}

As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated automatically by \rmv{the}
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 9 and 11, \ed{by}  \guido{This is vague:} \rmv{runtime checks.} 
\new{assuring that the annotated input dependencies {\tt line} and {\tt pattern}  are 
respected at runtime.}
\guido{Not clear. Need more details here. Adding:} \new{Pointer-based dependencies
can address various memory regions. Hence, in order to enable the automatic
synthesis of task annotations a system must be able to precisely define which 
memory region a pointer dependency is pointing to.}
\guido{Not clear why it is not important for correctness.}
Therefore, the ability to disambiguate memory regions statically is not
essential for correctness, although it is important for efficiency reasons,
as we will mention in Challenge~\ref{ch:Regions}.

\subsection{The Difficulties}
\label{sub:dif}

The automatic insertion of effective OpenMP annotations into C/C++ programs
required us to deal with a number of challenges.
If left unsolved, these challenges would restrict our interventions in programs
to trivial annotations, which would hardly be of any use.
Our first challenge is inherent to any automatic parallelization system.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.95\columnwidth]{images/ex_Regions}
\caption{Challenges~\ref{ch:Regions} and~\ref{ch:cost}: identifying memory regions
with symbolic limits, and using runtime information to estimate the profit of
tasks.}
\label{fig:ex_Regions}
\end{center}
\end{figure}

@@

Figure~\ref{fig:ex_Regions} illustrates Challenge~\ref{ch:Regions}, and shows
how we solve it\footnote{Figures~\ref{fig:ex_Regions}-\ref{fig:ex_cutoff} shows
programs that we have annotated without any human intervention.
Statements in black font are part of the original program.
Annotations appear in grey.}.
That program receives an $\mathsf{M}\times\mathsf{N}$ matrix \textsf{V}, in
linearized format, and produces a vector \textsf{U}, so that \textsf{U[i]}
contains the sum of all the elements in line \textsf{i} of matrix \textsf{V}.
For reasons to be considered in Section~\ref{sub:identification},
our static analysis determines that each iteration of the outermost loops
could be made into a task.
Thus, tasks comprise the innermost loop, and traverse the memory region
between addresses $\textsf{\&V + i * N}$ and $\textsf{\&V + i * N + M}$.
The identification of such ranges involves the use of a symbolic algebra,
which we have borrowed from the compiler-related literature, as we
explain in Section~\ref{sub:symb}.
Figure~\ref{fig:ex_Regions} also introduces the second challenge that we
tackle:

@@

The creation of tasks involves a heavy runtime cost due to allocation,
scheduling and real-time management of the task dependence graph.
Ideally, this cost should be paid only for tasks that perform an amount of work
sufficiently large to pay for their management.
Being an interesting program property, on Rice's sense~\cite{Rice53}, the amount
of work performed by a task cannot be discovered statically.
As we show in Section~\ref{sub:profit}, we can try to approximate this quantity, using,
to this end, program symbols, which are replaced with actual values at runtime.
For instance, in Figure~\ref{fig:ex_Regions}, we know that the body of the
innermost loop is formed by five instructions.
Thus, we approximate the amount of work performed by a task with the
expression \textsf{5 * M}.
We use the runtime value of \textsf{M} to determine, during the execution of the
program, if we create a task, or not.
Such test is carried out by the guard at line 7 of the figure, which is part of
OpenMP's syntax. Also, we provide a reliable estimate on the \emph{workload cutoff}
from which a task can be safely spawned without producing performance overhead.
This \emph{cutoff} considers factors such as number of available cores
and runtime information on the task dispatch cost in terms of machine
instructions. This is further explained in Section \ref{sub:profit}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.9\columnwidth]{images/ex_cutoff}
\caption{Challenge~\ref{ch:cutoff}: bounding the creation of recursive tasks with
counter associated with function calls.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\columnwidth]{images/alg_main}
\caption{The main steps of our code generator.}
\label{fig:alg_main}
\end{center}
\end{figure}

@@

The algorithm to find tasks is illustrated in Figure \ref{fig:alg_main}. It uses
information of two abstract program structures: the \emph{Control Flow Graph (CFG)} and
the \emph{Program Dependence Graph (PDG)}. The key intuition behind
finding potential tasks relies on a concept that we call \emph{Helices}. A \emph{Helix}
is a set of nodes in the PDG that depend on a Strong Connected Component (SCC) but \emph{do not
integrate the SCC}. Strong Connected Components in the Dependence Graph represent statically
detected iteration dependence, so it arises with loops or with recursive patterns (when the Graph is
interprocedural).  \Figure \ref{fig:helix}  shows what represents a \emph{Helix}.
The insight behind \emph{Helices} is that they represent instructions that are deemed
to repeat more than once and, since they do not have a cycle, they can be executed in parallel.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\columnwidth]{images/expand_alg}
\caption{Discovery of tasks via expansion of hammock regions.}
\label{fig:expand_alg}
\end{center}
\end{figure}

@@

\section{Evaluation}
\label{sec:eval}

%We aimed to show that our tool exceeded in three main aspects that surround parallel programming. Thus our work evaluation stands on these 3 pillars:

% In section 4. “Evaluation", subsection “Performance"; you state the only reason why programmers will strive for parallelism would be to achieve higher performance (faster program execution). I do not fully agree, and I believe we discussed that point earlier. There may be cases where performance of a single big core would be good enough to meet some target schedule (either response time or frame rate, etc.) still you may want to explore parallel versions of the code running on a cluster of smaller cores or a combination of a slow CPU and GPU to achieve better power efficiency.

@@

%\begin{itemize}

%\item Practicality
%\item Applicability
%\item Performance

%\end{itemize}

%\textbf{\textit{Practicality}}. Programming in a parallel pattern is deemed difficult by programmers mainly because of the complexity of its data structures and algorithms. We show that our tool is simple and practical, for it does not involve changing the code and data structures in order to fit in a parallel pattern. The code is simply annotated by compiler directives.

%\textbf{\textit{Applicability}}.

%\textbf{\textit{Performance}}. When it comes to parallel computing, it is undeniable that the main achievable goal should be performance. There's no reason why programmers would opt for parallelism other than the desire for faster programs. Therefore, we ought to show that the programs annotated with TaskMiner are faster than their serial counterparts.

@@

@@

>>>>

\section{Challenges}
\label{sec:ovf}

This paper presents a tool that annotates source C code with OpenMP task
directives.
This tool uses novel techniques to fulfill its purpose.
Such techniques solve four challenges, which Section~\ref{sub:dif}
illustrates with examples.
However, before we dive into these challenges, the reader must understand
the main benefits of using a runtime environment such as OpenMP's, for
such advantages have guided most of our designing decisions.
Section~\ref{sub:adv} introduces this discussion.

\subsection{The OpenMP Runtime System}
\label{sub:adv}

The main advantage of using OpenMP annotations to split a program into parallel
tasks is
the ability  of leaving the job of  handling dependences to a runtime.
The OpenMP runtime maintains a task dependence graph which dynamically exposes more parallelism 
opportunities than those resulting from a conservative  compile-time  analysis.
In particular, the runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its correctness~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.

Dependence tracking can be complex, involving checks over ranges of addresses
(if the runtime system treats dependencies across memory ranges).
It can also include memory labelling and renaming (if the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a dynamic topological ordering of
this graph~\cite{Planas15}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_book_filter}
\caption{The benefits of the OpenMP runtime environment.
Neither the runtime checks of Alves~\cite{Alves15} or Rus~\cite{Rus02}, nor
Whaley's context and flow sensitive alias analysis would be able to ensure the
correctness of the automatic parallelization of this program.}
\label{fig:ex_book_filter}
\end{center}
\end{figure}

The OpenMP runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
In such programs, control constructs like {\tt if} statements make the
execution  of certain pieces of code dependent on the program's input.
The runtime can capture such dependences, in contrast to static analysis tools.
As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated automatically by
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 9 and 11, by assuring that the annotated input dependencies
{\tt line} and {\tt pattern}  are 
respected at runtime.

\subsection{Challenges}
\label{sub:dif}

As seen in Section~\ref{sub:adv}, the OpenMP runtime liberates us
from the burden of having to track dependences between pointers statically.
However, the automatic insertion of effective task annotations into
programs still required us to deal with a number of challenges.
If left unsolved, these challenges would restrict our interventions
to trivial annotations -- hardly be of any use.
Our first challenge is inherent to any automatic parallelization system.

@@

\begin{figure}[b!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_Regions}
\caption{Challenges~\ref{ch:Regions} and~\ref{ch:cost}: identifying memory regions
with symbolic limits, and using runtime information to estimate the profit of
tasks.}
\label{fig:ex_Regions}
\end{center}
\end{figure}

Figure~\ref{fig:ex_Regions} illustrates Challenge~\ref{ch:Regions}, and shows
how we solve it\footnote{Whenever we show code, statements in black font are part
of the original program.
The annotations that we create automatically appear in grey.}.
That program receives an $\mathsf{M}\times\mathsf{N}$ matrix \textsf{V}, in
linearized format, and produces a vector \textsf{U}, so that \textsf{U[i]}
contains the sum of all the elements in line \textsf{i} of matrix \textsf{V}.
For reasons to be considered in Section~\ref{sub:identification},
our static analysis determines that each iteration of the outermost loops
could be made into a task.
Thus, tasks comprise the innermost loop, and traverse the memory region
between addresses $\textsf{\&V + i * N}$ and $\textsf{\&V + i * N + M}$.
The identification of such ranges involves the use of a symbolic algebra,
which we have borrowed from the compiler-related literature, as we
explain in Section~\ref{sub:symb}.
Figure~\ref{fig:ex_Regions} also introduces the second challenge that we
tackle:

@@

The creation of tasks involves a heavy runtime cost due to allocation,
scheduling and real-time management of the dependence graph
(see Sec.~\ref{sub:adv}).
Ideally, this cost should be paid only for tasks that perform an amount of work
sufficiently large to pay for their management.
Being an interesting program property, on Rice's sense~\cite{Rice53}, the amount
of work performed by a task cannot be discovered statically.
As we show in Section~\ref{sub:profit}, we can try to approximate this quantity, using,
to this end, program symbols, which are replaced with actual values at runtime.
For instance, in Figure~\ref{fig:ex_Regions}, we know that the body of the
innermost loop is formed by five instructions.
Thus, we approximate the amount of work performed by a task with the
expression \textsf{5 * M}.
We use the runtime value of \textsf{M} to determine, during the execution of the
program, if we create a task, or not.
Such test is carried out by the guard at line 7 of the figure, which is part of
OpenMP's syntax. Also, we provide a reliable estimate on the \emph{workload cutoff}
from which a task can be safely spawned without producing performance overhead.
This \emph{cutoff} considers factors such as number of available cores
and runtime information on the task dispatch cost in terms of machine
instructions. This is further explained in Section \ref{sub:profit}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_cutoff}
\caption{Challenge~\ref{ch:cutoff}: bounding the creation of recursive tasks with
counter associated with function calls.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/alg_main}
\caption{The main steps of our code generator.}
\label{fig:alg_main}
\end{center}
\end{figure}

@@

The algorithm to find tasks is illustrated in Figure \ref{fig:alg_main}. It uses
information of two abstract program structures: the \emph{Control Flow Graph (CFG)} and
the \emph{Program Dependence Graph (PDG)}. The key intuition behind
finding potential tasks relies on a concept that we call \emph{Helices}. A \emph{Helix}
is a set of nodes in the PDG that depend on a Strong Connected Component (SCC) but \emph{do not
integrate the SCC}. Strong Connected Components in the Dependence Graph represent statically
detected iteration dependence, so it arises with loops or with recursive patterns (when the Graph is
interprocedural).  Figure~\ref{fig:helix}  shows what represents a \emph{Helix}.
The insight behind \emph{Helices} is that they represent instructions that are deemed
to repeat more than once and, since they do not have a cycle, they can be executed in parallel.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/expand_alg}
\caption{Discovery of tasks via expansion of hammock regions.}
\label{fig:expand_alg}
\end{center}
\end{figure}

@@

\section{Evaluation}
\label{sec:eval}
% In section 4. “Evaluation", subsection “Performance"; you state the only reason why programmers will strive for parallelism would be to achieve higher performance (faster program execution). I do not fully agree, and I believe we discussed that point earlier. There may be cases where performance of a single big core would be good enough to meet some target schedule (either response time or frame rate, etc.) still you may want to explore parallel versions of the code running on a cluster of smaller cores or a combination of a slow CPU and GPU to achieve better power efficiency.

@@

We performed a thorough evaluation on \Taskminer{}'s applicability in real programs. Our test framework is divided in five aspects:

\begin{itemize}
\item Performance
\item Optimizations
\item Versatility
\item Applicability
\item Practicality
\end{itemize}

%PERFORMANCE
%Here, we show the results for our small benchmarks (toys & bots)
\textbf{\textit{Performance}}.

@@

%OPTIMIZATIONS
%Here, we show the results on two big optimizations: recursion depth and cost model.
%We can use the same benchmarks as above.
\textbf{\textit{Optimizations}}.

@@

%VERSATILITY
%The goal here is to show that TaskMiner is a versatile tool, capable of finding many types of task parallelism in the code. 
%And the types of tasks that are going to be mined can be easily set, should the programmer ever desires to.
\textbf{\textit{Versatility}}.

%APPLICABILITY
%We'll show TaskMiner application on larger benchmarks, such as those of spec or those of the LLVM's test-suite.
\textbf{\textit{Versatility}}.

%PRACTICALITY
%The main objective here is to show that TaskMiner can be easily utilized by a programmer to find tasks in their source code. 
%We'll test TaskMiner's running time.
\textbf{\textit{Practicality}}.
